---
title: "IHA_key"
author: "Justin Pomeranz"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(RcppRoll)
post <- read_csv(here::here("data/IHA/post_data.csv"))
source(here::here("iha_pre_source_for_key.R"))
```

# Introduction  

This document is the key for the Indicators of Hydrologic Alteration (IHA) assignment given to ENVS433- Aquatic Restoration Students at Colorado Mesa University


## Plot

```{r}
ggplot(post, 
       aes(x = date, 
           y = cfs)) +
  geom_line() +
  theme_bw() +
  labs(title = "Mean Daily CFS values at Lee's Ferry, 1990-1999",
       y = "cubic feet per second")
```

### Group 1
```{r}
group1_post <- post %>%
  group_by(m) %>%
  summarize(m_mean = mean(cfs))
# print out your new object to see it in the console
group1_post
```

### Group 2
#### 1-day values  
```{r}
post_g2_1d <- post %>%
  group_by(y) %>%
  summarise(min_1 = min(cfs), 
            max_1 = max(cfs)) %>%
  summarise(m_min_1 = mean(min_1), 
            m_max_1 = mean(max_1))
```

#### 3-day
```{r}
# 3-day min and max
post_g2_3d <- post %>%
  group_by(y) %>%
  reframe(mean_3 = roll_mean(cfs, n = 3)) %>%
  group_by(y) %>%
  summarise(min_3 = min(mean_3),
            max_3 = max(mean_3)) %>%
summarise(m_min_3 = mean(min_3),
          m_max_3 = mean(max_3))

post_g2_3d
```


#### 7-day

```{r}
# 7-day min and max
post_g2_7d <- post %>%
  group_by(y) %>%
  reframe(mean_7 = roll_mean(cfs, n = 7)) %>%
  group_by(y) %>%
  summarise(min_7 = min(mean_7),
            max_7 = max(mean_7)) %>%
summarise(m_min_7 = mean(min_7),
          m_max_7 = mean(max_7))

post_g2_7d
```

#### 30-day  

* 30-day 

```{r}
# 30-day min and max
post_g2_30d <- post %>%
  group_by(y) %>%
  reframe(mean_30 = roll_mean(cfs, n = 30)) %>%
  group_by(y) %>%
  summarise(min_30 = min(mean_30),
            max_30 = max(mean_30)) %>%
summarise(m_min_30 = mean(min_30),
          m_max_30 = mean(max_30))

post_g2_30d
```

#### 90-day 


* 90-day 

```{r}
# 90-day min and max
post_g2_90d <- post %>%
  group_by(y) %>%
  reframe(mean_90 = roll_mean(cfs, n = 90)) %>%
  group_by(y) %>%
  summarise(min_90 = min(mean_90),
            max_90 = max(mean_90)) %>%
summarise(m_min_90 = mean(min_90),
          m_max_90 = mean(max_90))

post_g2_90d
```

```{r}
# combine all the columns of results for the 5 windows into one object
post_g2 <- bind_cols(post_g2_1d,
          post_g2_3d,
          post_g2_7d,
          post_g2_30d,
          post_g2_90d) %>%
  pivot_longer(cols = starts_with("m_"), names_to = "metric", values_to = "mean") %>%
  separate(metric, c("drop", "metric", "day-window")) %>%
  select(!drop) 
```



### Group 3

```{r}
post_min_julian <- post %>%
  group_by(y) %>%
  filter(cfs == min(cfs)) %>%
  select(cfs, y, j_date) %>%
  ungroup() %>%
  summarise(min_j_date = mean(j_date))

post_min_julian
```

* calculate for max flow  

* similar code as above, but note name changes and function inside filter to `max` 

```{r}
post_max_julian <- post %>%
  group_by(y) %>%
  filter(cfs == max(cfs)) %>%
  select(cfs, y, j_date) %>%
  ungroup() %>%
  summarise(max_j_date = mean(j_date))

post_max_julian
```

# DM and %D  

## Group 1 DM
```{r}
# deviation magnitude
g1_dm <- group1_post - group1_pre
g1_dm
```

## Group 1 %D
```{r}
g1_pd <- g1_dm / group1_pre * 100
g1_pd
```

## Group 2 DM
```{r}
# deviation magnitude
g2_dm <- post_g2$mean - pre_g2$mean_cfs
data.frame(metric = post_g2$metric,
           "day_window" = post_g2$`day-window`,
           dm = g2_dm)
```

## Group 2 %D
```{r}
g2_pd <- g2_dm / pre_g2$mean_cfs * 100
data.frame(metric = post_g2$metric,
           "day_window" = post_g2$`day-window`,
           "percent_dm" = g2_pd)

```

## group 3 Min Julian DM
```{r}
g3_min_julian_dm <- post_min_julian - pre_min_julian
g3_min_julian_dm
```
## group 3 Min Julian %D
```{r}
g3_min_julian_dm / pre_min_julian * 100
```


## group 3 Max Julian DM
```{r}
g3_max_julian_dm <- post_max_julian - pre_max_julian
g3_max_julian_dm
```
## group 3 Max Julian %D
```{r}
g3_max_julian_dm / pre_max_julian * 100
```

