---
title: "IHA-assignment"
author: "Justin Pomeranz"
date: "7/14/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(RcppRoll)
```

# Overview  

This assignment will explore how river regulation alters the natural flow regime. You will use real data from the Colorado River at Lee's Ferry before and after Glen Canyon dam was built. 

# Introduction  

Rivers are subjected to variation in flow conditions across time. Within a given year, they can be characterized by a natural flow regime (Poff et al. 1990), with predictable and consistent periods of high and low flow conditions. 

Dams are built on rivers around the world for numerous reasons including water storage, flood control, and hydropower operations, all of which can alter the natural flow regime. 

## Glen Canyon Dam  

Glen Canyon Dam was built on the Colorado River to generate hydropower and supply municipal and agricultural water to the desert southwest region of the United States. Completed in 1965, the 710-foot tall dam forms Lake Powell, which can store > 25 million acre-feet of water and is one of the largest human-made reservoirs in the US. The reservoir was filled to capacity in 1980, but the level has fluctuated dramatically since then, particularly in recent decades in response to a ~20 year drought. 

## Alterations to flow  

```{r echo=FALSE, message=FALSE, warning=FALSE}
df <- read_csv(here::here("data/IHA/cfs-1930-2023.csv"))

df %>%
  # mutate(m = month(date),
  #        y = year(date)) %>%
  # group_by(y, m) %>%
  # reframe(month_mean = mean(cfs)) %>%
  ggplot(aes(x = date, y = cfs)) +
  geom_line() +
  theme_bw() +
  labs(caption = str_wrap("Figure 1. Flow data in cubic feet per second (cfs) for Lee's Ferry below Glen Canyon Dam. Note the extremely low flows beginning in the mid 1960's when the dam was closed, and reduced variabiliaty after ~ 1980 when the reservoir was full."))
```

The closing of Glen Canyon Dam dramatically altered the flow at Lee's Ferry, as seen in figure 1. Qualitatively, we can see much higher variation in the magnitudes of high- and low-flows. 

# Assignment  

This assignment will be completed in groups of 2-3 and has four main parts:  
1. Calculate IHA metrics (groups 1 - 3, Richter et al. 1996) For Lee's ferry before and after the dam was closed.  
2. Use the Range of Variability Approach (RVA, Richter et al. 1997) for selected metrics from group 2 to make management recommendations for Glen Canyon Dam operations.  
3. Compare your recommendations with flow conditions from experimental high flows in 2023, and experimental low-flows in 2020.  
4. You will write up your analysis to summarize and discuss points 1-3.  

## Quantifying the IHA and RVA  

In this assignment, you will calculate IHA groups 1-3 for both pre- (1930-1939) and post-dam closure data (1990-1999). 

The data are available on D2L (`pre_data.csv` and `post_data.csv`, respectively). A tutorial is also available on D2L (**IHA and RVA tutorial**) explaining how to calculate the IHA metrics for the pre-data as well as the RVA approach (points 1 and 2 above). We will also go over this tutorial in class. 

You will be responsible for making these calculations on the post-data on your own. 

## Comparing with experimental flows  

In the last few years, there have been experimental high- and low-flow releases from Glen Canyon Dam to investigate whether or not these flows had ecological benefits. You can read more about the [experimental flood in 2023](https://www.nps.gov/grca/learn/nature/hfe-23.htm), as well as the [experimental low, Bug-flows in 2020](https://www.usbr.gov/newsroom/newsroomold/newsrelease/detail.cfm?RecordID=70708). Daily mean flow data for both of these experiments are available on D2L as `high-flow-2023.csv` and `bug_flows.csv`. 

## Data sources  

The data provided for this assignment was downloaded from the USGS website, gauge number 09380000. The downloaded data were curated, cleaned and modified. 

If you would like to access the raw, daily data, you can go to the [USGS water data center](https://waterdata.usgs.gov/nwis/dv/?site_no=09380000&referred_module=sw) to download tab-separated data files. You can also view provisional data at the [USGS monitoring site](https://waterdata.usgs.gov/monitoring-location/09380000/#parameterCode=00060&startDT=1920-04-01&endDT=2023-05-01). 

If you would like help on how to access, download, and modify the data to be read in to R, please get in touch with me or come to my office hours. 