---
title: "Plotting_RVA_tutorial"
author: "Justin Pomeranz"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(RcppRoll)
pre <- read_csv(here::here("data/IHA/pre_data.csv"))
bugs <- read_csv(here::here("data/IHA/bug_flows.csv"))
```

# Overview  

This tutorial will show you how to plot the range of variability approach (RVA) for the minimum 1-day flow and compare it with an experimental low flow from Glen Canyon Dam. 

# Preliminaries  

* Open RStudio  

  * Make sure you are in your `ENVS433` project (look in the top right area of RStudio)  
  
* Make sure you have a folder called `data` in your project and that the `pre_data.csv` file is in that folder.  

* load the packages you need by running the following code:
```{r, eval=FALSE}
library(tidyverse)
library(RcppRoll)
```

  * If there was an error, make sure you have the `tidyverse` and `RcppRoll` packages downloaded onto your machine  
  
  * Remember you only need to install the packages once per machine, but you need to load them using the `library()` command every time you open RStudio.  
  
  * If needed, you can install the packages by running the following code: `install.packages("tidyverse")`, and `install.packages("RcppRoll")`, respectively. Make sure to re-run the `library()` commands above if needed.  
  
  
## Load the data into R  

If you have the correct file in your `data` folder, you can load it into R using the following command:
```{r, eval=FALSE}
pre <- read_csv("data/pre_data.csv")
bugs <- read_csv("data/bug_flows.csv")
```

* Check that the data was loaded correctly.  


# need to fix date in bugs data

1,565 and 3,352 CFS
```{r}
bugs %>%
  ggplot(aes(x = date, 
             y = cfs)) +
  geom_line()
```

Note that the low-flows in May-June were above 10,000 CFS, and once the experiment began the low-flows were ~9,000 cfs (at least at first). 

### Group 2  

* We will only be looking at the 1-day minimums in this tutorial.  


#### 1-day values  

Recall that the RVA for the 1-day minimums was calculated as follows: 

```{r}
pre %>%
  group_by(y) %>%
  summarise(min_1 = min(cfs)) %>%
  summarise(m_min_1 = mean(min_1), 
            # add sd() command here
            sd_min_1 = sd(min_1)) %>%
  pivot_longer(1:2, 
               names_to = "metric", 
               values_to = "value") %>%
  separate(1, c("stat", "metric", "day_window")) %>%
  pivot_wider(names_from = stat) %>%
  mutate(rva_low = m - sd,
         rva_high = m + sd) %>%
  select(metric, rva_low, rva_high)
```

This means that the RVA interval for the 1-day minimum flow is between 1,565 and 3,352 CFS. 

We can now add reference lines to our original plot, and compare the RVA with what was actually observed. 

```{r}
bugs %>%
  ggplot(aes(x = date, 
             y = cfs)) +
  geom_line() +
  geom_hline(aes(yintercept = 1565), 
               linetype = "dashed") + # low interval
    geom_hline(aes(yintercept = 3352), 
               linetype = "dashed") 
```

Here, we can see that the experimental low flows were much higher than the RVA for the 1-day low flows. 


### Hypothetical result  

Just for comparison, and so you know what it looks like when our experimental flows *are* within the recommended RVA, here's the plot of a hypothetical result. I'm also changing the `linetype` for illustrative purposes. 

```{r}
bugs %>%
  ggplot(aes(x = date, 
             y = cfs)) +
  geom_line() +
  geom_hline(aes(yintercept = 8000), 
               linetype = "twodash") + # low interval
    geom_hline(aes(yintercept = 9500), 
               linetype = "twodash") 
```
Here, we can see that the first four low flows fall within the interval. It looks like the 5th-7th low flow also falls within the region, although we may need to "zoom in" to tell for sure.  

#### "Zooming in"

```{r}
bugs %>%
  ggplot(aes(x = date, 
             y = cfs)) +
  geom_line() +
  geom_hline(aes(yintercept = 8000), 
               linetype = "twodash") + # low interval
    geom_hline(aes(yintercept = 9500), 
               linetype = "twodash") +
  coord_cartesian(ylim = c(8000, 10000))
```
It does appear that the 1st to 7th low-flows do fall within the RVA interval. 

# Moving forward    

## Complete part 4  

### Minimum flows with `bugs` data  

You will need to modify the code above to plot the RVA for the minimum flow in the 3- and 7-day windows. 


### Maximum flows with `high-flow-2023.csv` data  

You will need to modify the code above to plot the RVA for the minimum flow in the 1-, 3-, and 7-day windows. Instead of using the same `bugs` data as above, you will need to compare it to the `high-flow-2023.csv`, which you can download from D2L. 




