---
title: "RVA_KEY"
author: "Justin Pomeranz"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(RcppRoll)
pre <- read_csv(here::here("data/IHA/pre_data.csv"))
bugs <- read_csv(here::here("data/IHA/bug_flows.csv"))
high <- read_csv(here::here("data/IHA/high-flow-2023.csv"))
```

# Overview  

This is the key for the RVA section of the assignment

# RVA calculations

## Group 2  

### 1-day values  

We can modify this code slighty to also calculate the SD for the min and max values. 

```{r}
pre_g2_1d_sd <- pre %>%
  group_by(y) %>%
  summarise(min_1 = min(cfs), 
            max_1 = max(cfs)) %>%
  summarise(m_min_1 = mean(min_1),
            sd_min_1 = sd(min_1),
            m_max_1 = mean(max_1),
            sd_max_1 = sd(max_1)) %>%
  pivot_longer(cols = 1:4, names_to = "metric", values_to = "value") %>%
  separate(col = 1, c("summary_stat", "range", "day-window")) %>%
  pivot_wider(names_from = 1) %>%
  mutate(low_sd = m - sd,
         hi_sd = m+sd) %>%
  select(-m, -sd)

pre_g2_1d_sd
```

### 1-day plots

#### Min, compared with bugs
```{r}
ggplot(bugs, 
       aes(x = date, 
           y = cfs)) +
  geom_line() +
  theme_bw() +
  labs(title = "Mean Daily CFS values at Lee's Ferry, BUGS-flows",
       y = "cubic feet per second",
       caption = "The RVA for 1-day minimums is shown with a black horizontal line (mean - sd) and a red dashed line (mean + sd)") +
  geom_hline(yintercept = 1566) +
  geom_hline(yintercept = 3352,
             color = "red", 
             linetype = "dashed") +
  coord_cartesian(ylim = c(1500, 20000))
```

#### Max, compared with high flow
```{r}
ggplot(high, 
       aes(x = date, 
           y = cfs)) +
  geom_line() +
  theme_bw() +
  labs(title = "Mean Daily CFS values at Lee's Ferry, experimental high flows 2023",
       y = "cubic feet per second",
        caption = "The RVA for 1-day maximums is shown with a black horizontal line (mean - sd) and a red dashed line (mean + sd)") +
   geom_hline(yintercept = 43870) +
   geom_hline(yintercept = 99770,
              color = "red", 
              linetype = "dashed") +
   coord_cartesian(ylim = c(10000, 100000))
```


#### 3-day values  

* This code is copied from the IHA tutorial

```{r}
# 3-day min and max
pre_g2_3d_sd <- pre %>%
  group_by(y) %>%
  reframe(mean_3 = roll_mean(cfs, n = 3)) %>%
  group_by(y) %>%
  summarise(min_3 = min(mean_3),
            max_3 = max(mean_3)) %>%
  summarise(m_min_3 = mean(min_3),
          sd_min_3 = sd(min_3),
          m_max_3 = mean(max_3), 
          sd_max_3 = sd(max_3)) %>%
  pivot_longer(cols = 1:4, names_to = "metric", values_to = "value") %>%
  separate(1, c("summary_stat", "range", "day-window")) %>%
  pivot_wider(names_from = 1) %>%
  mutate(low_sd = m - sd,
         hi_sd = m+sd) %>%
  select(-m, -sd)

pre_g2_3d_sd
```

### 3-day plots

#### Min, compared with bugs
```{r}
ggplot(bugs, 
       aes(x = date, 
           y = cfs)) +
  geom_line() +
  theme_bw() +
  labs(title = "Mean Daily CFS values at Lee's Ferry, BUGS-flows",
       y = "cubic feet per second",
       caption = "The RVA for 3-day minimums is shown with a black horizontal line (mean - sd) and a red dashed line (mean + sd)") +
  geom_hline(yintercept = 1758) +
  geom_hline(yintercept = 3490,
             color = "red", 
             linetype = "dashed") +
  coord_cartesian(ylim = c(1500, 20000))
```

#### Max, compared with high flow
```{r}
ggplot(high, 
       aes(x = date, 
           y = cfs)) +
  geom_line() +
  theme_bw() +
  labs(title = "Mean Daily CFS values at Lee's Ferry, experimental high flows 2023",
       y = "cubic feet per second",
        caption = "The RVA for 3-day maximums is shown with a black horizontal line (mean - sd) and a red dashed line (mean + sd)") +
   geom_hline(yintercept = 43870) +
   geom_hline(yintercept = 99770,
              color = "red", 
              linetype = "dashed") +
   coord_cartesian(ylim = c(10000, 100000))
```

#### 7-day values

We can repeat the process for the 7-day values. 

* calculations copied from the IHA tutorial, with the sd calculations added

```{r}
# 7-day min and max
pre_g2_7d_sd <- pre %>%
  group_by(y) %>%
  reframe(mean_7 = roll_mean(cfs, n = 7)) %>%
  group_by(y) %>%
  summarise(min_7 = min(mean_7),
            max_7 = max(mean_7)) %>%
summarise(m_min_7 = mean(min_7),
          sd_min_7 = sd(min_7),
          m_max_7 = mean(max_7), 
          sd_max_7 = sd(max_7)) %>%
  pivot_longer(cols = 1:4, names_to = "metric", values_to = "value") %>%
  separate(1, c("summary_stat", "range", "day-window")) %>%
  pivot_wider(names_from = 1) %>%
  mutate(low_sd = m - sd,
         hi_sd = m+sd) %>%
  select(-m, -sd)
  

pre_g2_7d_sd
```
### 7-day plots

#### Min, compared with bugs
```{r}
ggplot(bugs, 
       aes(x = date, 
           y = cfs)) +
  geom_line() +
  theme_bw() +
  labs(title = "Mean Daily CFS values at Lee's Ferry, BUGS-flows",
       y = "cubic feet per second",
       caption = "The RVA for 7-day minimums is shown with a black horizontal line (mean - sd) and a red dashed line (mean + sd)") +
  geom_hline(yintercept = 1923) +
  geom_hline(yintercept = 3741,
             color = "red", 
             linetype = "dashed") +
  coord_cartesian(ylim = c(1500, 20000))
```

#### Max, compared with high flow
```{r}
ggplot(high, 
       aes(x = date, 
           y = cfs)) +
  geom_line() +
  theme_bw() +
  labs(title = "Mean Daily CFS values at Lee's Ferry, experimental high flows 2023",
       y = "cubic feet per second",
        caption = "The RVA for 7-day maximums is shown with a black horizontal line (mean - sd) and a red dashed line (mean + sd)") +
   geom_hline(yintercept = 40439) +
   geom_hline(yintercept = 92999,
              color = "red", 
              linetype = "dashed") +
   coord_cartesian(ylim = c(10000, 100000))
```

